ATAC and RNA correlation

## imports
```{r}
library(tidyverse)
library(data.table)
library(viridis)
library(ggrastr)
library(motifmatchr)
library(ggpointdensity)
```

## Correlate by TF group vs GFP group
```{r}
meta <- read_tsv("../03-rna/rna_sample_sheet.txt")

celltype <- "HEK293T"
cond2 <- "GFP"
log2fc.cutoff <- 0.58
padj.cutoff <- 0.05

atacdiff_dir <- "../../output/02-atac/04/deseq_data_TF/"
rnadiff_dir <- paste0("../../output/03-rna/01/tf_group/", celltype)

plotdir <- "../../plots/02-atac/05/tf_group"
outdir <- "../../output/02-atac/05/tf_group"

dir.create(plotdir, showWarnings = F, recursive = T)  
dir.create(outdir, showWarnings = F, recursive = T)

submeta <- meta[meta$CellType==celltype,]
cond1_ls <- unique(submeta$TF)
cond1_ls <- cond1_ls[cond1_ls != cond2]
```

```{r}
cordf <- data.frame(row.names=cond1_ls, n_atac_diff_peaks=rep(0, length(cond1_ls)), n_rna_diff_genes=rep(0, length(cond1_ls)))

for (cond1 in cond1_ls){
  if (cond1 %in% c("EF1a-KLF1", "EF1a-SPI1", "tet-CTCF")){
    namechange <- list("EF1a-KLF1"="KLF1", "EF1a-SPI1"="SPI1", "tet-CTCF"="tetCTCF")
    ataccond1 <- namechange[cond1]
  } else{
    ataccond1 <- cond1
  }
  atacdiff <- fread(paste0(atacdiff_dir, "/deseq_", ataccond1, "_vs_", cond2, "_filteredConsensus_full.tsv"))
  rnadiff <- readRDS(paste0(rnadiff_dir, "/deseq_dds_", gsub("-", ".", cond1), "_vs_", cond2, ".rds"))
  
  atacdiff_sig <- atacdiff %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  rnadiff_sig <- rnadiff %>% as.data.frame %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  
  cordf[cond1, "n_atac_diff_peaks"] <- nrow(atacdiff_sig)
  cordf[cond1, "n_atac_diff_peaks_up"] <- nrow(atacdiff_sig %>% dplyr::filter(log2FoldChange>0))
  cordf[cond1, "n_atac_diff_peaks_down"] <- nrow(atacdiff_sig %>% dplyr::filter(log2FoldChange<0))
  cordf[cond1, "n_rna_diff_genes"] <- nrow(rnadiff_sig)
  cordf[cond1, "n_rna_diff_genes_up"] <- nrow(rnadiff_sig %>% dplyr::filter(log2FoldChange>0))
  cordf[cond1, "n_rna_diff_genes_down"] <- nrow(rnadiff_sig %>% dplyr::filter(log2FoldChange<0))
}

colpal <- read_tsv("../utils/colpal_TF.tsv") %>% 
  column_to_rownames(var="TF")

cordf$TF <- row.names(cordf)
cordf[cordf$TF=="EF1a-KLF1", "TF"] <- "KLF1"
cordf[cordf$TF=="EF1a-SPI1", "TF"] <- "SPI1"
cordf[cordf$TF=="tet-CTCF", "TF"] <- "tetCTCF"
cordf$TF <- as.factor(cordf$TF)
BL_stallion <- colpal[levels(cordf$TF),]
  
ggplot(cordf, aes(x=n_atac_diff_peaks, y=n_rna_diff_genes, col=TF, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) + ggrepel::geom_label_repel(show.legend = F)
ggsave(paste0(plotdir, "/atac_rna_ndiff_cor.pdf"), height=5, width=7)

ggplot(cordf, aes(x=log10(n_atac_diff_peaks), y=log10(n_rna_diff_genes), col=TF, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) + ggrepel::geom_label_repel(show.legend = F, max.overlaps = 15)
ggsave(paste0(plotdir, "/atac_rna_ndiff_cor_log10.pdf"), height=5, width=7)

ggplot(cordf[cordf$TF!="tetCTCF",], aes(x=log10(n_atac_diff_peaks), y=log10(n_rna_diff_genes), col=TF, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) + ggrepel::geom_label_repel(show.legend = F, max.overlaps = 15)
ggsave(paste0(plotdir, "/atac_rna_ndiff_cor_log10_notetctcf.pdf"), height=5, width=7)


write.table(cordf, paste0(outdir, "/atac_rna_ndiff_cor.txt"), sep="\t", quote=F)

cordf <- read.table(paste0(outdir, "/atac_rna_ndiff_cor.txt"), sep="\t")
```
### Plot a stacked bar plot to show TF group diff peaks/genes
```{r}
df_long <- tidyr::gather(cordf, key="atac_diffpeak_type", value="count", n_atac_diff_peaks_up, n_atac_diff_peaks_down)
df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
                                               
# Adjust the 'down' values to be negative
df_long$count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"] <- -df_long$count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"]

# Plot using ggplot
ggplot(df_long, aes(x=TF, y=count, fill=atac_diffpeak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="ATAC Differential Peaks (Up and Down)",
       y="Number of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/atac_ndiff_cor_updown_stacked_bar.pdf"), height=5, width=7)

# rna 
df_long <- tidyr::gather(cordf, key="rna_diffgene_type", value="count", n_rna_diff_genes_up, n_rna_diff_genes_down)
df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::arrange(desc(n_rna_diff_genes_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
                                               
# Adjust the 'down' values to be negative
df_long$count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"] <- -df_long$count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"]

# Plot using ggplot
ggplot(df_long, aes(x=TF, y=count, fill=rna_diffgene_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="RNA Differential Genes (Up and Down)",
       y="Number of Differential Genes", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/rna_ndiff_cor_updown_stacked_bar.pdf"), height=5, width=7)
```

### Correlation between ATAC log2fc and RNA log2fc (nearest gene) by group
```{r}
for (cond1 in cond1_ls){
  message(cond1)
  if (length(grep("EF1a|tet", cond1))==1){
    ataccond1 <- gsub("EF1a-", "", cond1) %>% gsub("tet-", "tet", .)
  } else{
    ataccond1 <- cond1
  }
  atacdiff <- fread(Sys.glob(paste0(atacdiff_dir, "/deseq_*", ataccond1, "_vs_*", cond2, "_filteredConsensus_full.tsv")))
  rnadiff <- readRDS(paste0(rnadiff_dir, "/deseq_dds_", gsub("-", ".", cond1), "_vs_", cond2, ".rds"))
  
  atacdiff_sig <- atacdiff %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  rnadiff_sig <- rnadiff %>% as.data.frame %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  
  # find the nearest gene of each differential peak
  atacdiff_sig <- atacdiff_sig %>% na.omit %>% dplyr::filter(gene_name %in% rnadiff$gene_name) # filter out any peaks with no matching nearest gene names in RNA data
  
  # collapse duplicate gene names in RNA table, sort by smallest padj then max log2fc then select the top match
  rnadiff_filt <- rnadiff %>% na.omit %>% as.data.frame %>% dplyr::filter(gene_name!="") %>% 
                  dplyr::group_by(gene_name) %>% dplyr::arrange(padj, desc(log2FoldChange)) %>% dplyr::slice(1)
  colnames(rnadiff_filt) <- paste0("rna_", colnames(rnadiff_filt))
  cat <- merge(atacdiff_sig, rnadiff_filt, all.x=T, by.x="gene_name", by.y="rna_gene_name")
  cat$is_deg <- cat$gene_name %in% rnadiff_sig$gene_name
  print(paste0(sum(cat$is_deg)/nrow(cat) * 100, " % of nearest genes near diff peaks are also diff genes"))
  
  write.table(cat, paste0(outdir, "/atac_rna_log2fc_cor_",cond1, "_vs_", cond2,".txt"), sep="\t", quote=F)

  p1 <- ggplot(cat, aes(x=log2FoldChange, y=rna_log2FoldChange)) + rasterise(geom_point(size=0.3)) + 
    xlab("Differential Peak ATAC Log2FC") + ylab("Nearest Gene RNA Log2FC ") + ggtitle(paste0(cond1, " vs ", cond2, "(n=", nrow(cat), ")"))
  ggsave(paste0(plotdir, "/atac_rna_log2fc_cor_", cond1, "_vs_", cond2, ".pdf"), height=5, width=5)
  
}


```
### Correlation between RNA log2fc and ATAC log2fc (peaks within 10kb) by group

```{r}
for (cond1 in cond1_ls){
  message(cond1)
  if (length(grep("EF1a|tet", cond1))==1){
    ataccond1 <- gsub("EF1a-", "", cond1) %>% gsub("tet-", "tet", .)
  } else{
    ataccond1 <- cond1
  }
  atacdiff <- fread(Sys.glob(paste0(atacdiff_dir, "/deseq_*", ataccond1, "_vs_*", cond2, "_filteredConsensus_full.tsv")))
  rnadiff <- readRDS(paste0(rnadiff_dir, "/deseq_dds_", gsub("-", ".", cond1), "_vs_", cond2, ".rds"))

  atacdiff_sig <- atacdiff %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  rnadiff_sig <- rnadiff %>% as.data.frame %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)

  # find all peaks within 10kb of each differential gene
  peaks_nearby <- atacdiff  %>% dplyr::filter((gene_name %in% rnadiff_sig$gene_name) & dist_to_tss<10000)

  rnadiff_filt <- rnadiff_sig[rnadiff_sig$gene_name!="",]
  colnames(rnadiff_filt) <- paste0("rna_", colnames(rnadiff_filt))
  cat <- merge(peaks_nearby, rnadiff_filt, all.y=T, by.x="gene_name", by.y="rna_gene_name")
  cat$is_deg <- cat$name %in% atacdiff_sig$name
  print(paste0(sum(cat$is_deg)/nrow(cat) * 100, " % of peaks within 10kb of diff genes are also diff peaks"))

  write.table(cat, paste0(outdir, "/atac_rna_log2fc_cor_peakswithin10kbdiffgenes_",cond1, "_vs_", cond2,".txt"), sep="\t", quote=F)
  
  # cat <- read.table(paste0(outdir, "/atac_rna_log2fc_cor_peakswithin10kbdiffgenes_",cond1, "_vs_", cond2,".txt"), sep="\t")
  p1 <- ggplot(cat, aes(x=rna_log2FoldChange, y=log2FoldChange)) + rasterise(geom_point(size=0.3)) +
    xlab("Differential Gene RNA Log2FC") + ylab("Nearby <10kb ATAC peaks Log2FC ") + ggtitle(paste0(cond1, " vs ", cond2, "(npeaks=", nrow(cat), ")"))
  p2 <- ggplot(cat, aes(x=rna_log2FoldChange, y=log2FoldChange)) + rasterise(geom_pointdensity(size=0.3)) +
    xlab("Differential Gene RNA Log2FC") + ylab("Nearby <10kb ATAC peaks Log2FC ") + ggtitle(paste0(cond1, " vs ", cond2, "(npeaks=", nrow(cat), ")")) + scale_color_viridis()
  pdf(paste0(plotdir, "/atac_rna_log2fc_cor_peakswithin10kbdiffgenes_", cond1, "_vs_", cond2, ".pdf"), height=5, width=5)
  print(p1)
  print(p2)
  dev.off()

  # try summarize all peaks nearby each gene into a gene score
  # formula from https://www.archrproject.com/bookdown/calculating-gene-scores-in-archr.html
  cat$weighted_atac_log2fc <- (exp(-abs(cat$dist_to_tss)/5000) + exp(-1)) * cat$log2FoldChange
  tmp <- cat %>% dplyr::group_by(gene_name) %>% dplyr::summarise(avg_weighted_atac_log2fc=mean(weighted_atac_log2fc), rna_log2FoldChange=mean(rna_log2FoldChange))
  p3 <- ggplot(tmp, aes(x=rna_log2FoldChange, y=avg_weighted_atac_log2fc)) + rasterise(geom_point(size=0.3)) + 
    xlab("Differential Gene RNA Log2FC") + ylab("Nearby <10kb ATAC peaks Mean Weighted Log2FC ") + ggtitle(paste0(cond1, " vs ", cond2, "(ngenes=", nrow(tmp), ")"))
  p4 <- ggplot(tmp, aes(x=rna_log2FoldChange, y=avg_weighted_atac_log2fc)) + rasterise(geom_pointdensity(size=0.3)) + 
    xlab("Differential Gene RNA Log2FC") + ylab("Nearby <10kb ATAC peaks Mean Weighted Log2FC ") + ggtitle(paste0(cond1, " vs ", cond2, "(ngenes=", nrow(tmp), ")")) + scale_color_viridis()
  pdf(paste0(plotdir, "/atac_rna_log2fc_cor_peakswithin10kbdiffgenes_weighted_", cond1, "_vs_", cond2, ".pdf"), height=5, width=5)
  print(p3)
  print(p4)
  dev.off()
}


```
### % of diff peaks up/down that contain motif of interest
```{r}
motifs <- readRDS("../../output/02-atac/03/motif.scores.JASPAR2020.rds")
motif_code_ls <- c(
                   "ALX4"="MA0681.2_PHOX2B",
                   "ELF1"="MA0473.3_ELF1",
                   "IRF4"="MA1419.1_IRF4",
                   "KLF1"="MA0039.4_KLF4",
                   "KLF4"="MA0039.4_KLF4",
                   "LEF1"="MA0768.1_LEF1",
                   "NR4A1"="MA1112.2_NR4A1",
                   "OCT4"="MA1115.1_POU5F1",
                   "PRDM1"="MA0508.3_PRDM1",
                   "SOX2"="MA0143.4_SOX2",
                   "SP4"="MA0685.1_SP4",
                   "SPI1"="MA0080.5_SPI1",
                   "TCF3"="MA0522.3_TCF3",
                   "MYC"="MA0059.1_MAX::MYC",
                   "MXD1"="MA0059.1_MAX::MYC",
                   "FOXO1"="MA0481.3_FOXP1", # no human FOXO1 motif
                   "FOXP1"="MA0481.3_FOXP1",
                   "FOXP3"="MA0850.1_FOXP3",
                   "GATA1"="MA0140.2_GATA1::TAL1",
                   "BACH2"="MA1101.2_BACH2",
                   "IKZF2"="MA1508.1_IKZF1",
                   "XBP1"="MA0844.1_XBP1")

cond1_ls <- cond1_ls[grep(paste0(names(motif_code_ls), collapse="|"), cond1_ls)]

out <- list()
for (cond1 in cond1_ls){
  message(cond1)
  if (length(grep("EF1a|tet", cond1))==1){
    ataccond1 <- gsub("EF1a-", "", cond1) %>% gsub("tet-", "tet", .)
  } else{
    ataccond1 <- cond1
  }
  submotifs <- motifs[,motif_code_ls[ataccond1]]
  submotifs <- submotifs[motifMatches(submotifs) %>% as.list %>% unlist ==TRUE,]
  
  df <- fread(Sys.glob(paste0(atacdiff_dir, "/deseq_*", ataccond1, "_vs_*", cond2, "_filteredConsensus_full.tsv")))
  up <- df %>% dplyr::filter(padj<=padj.cutoff & log2FoldChange>=log2fc.cutoff)
  down <- df %>% dplyr::filter(padj<=padj.cutoff & log2FoldChange<=-log2fc.cutoff)
  out[[cond1]] <- data.frame(npeak=c(nrow(up), nrow(down)),
            npeak_contain_motif_of_interest=c(sum(up$name %in% rownames(submotifs)), sum(down$name %in% rownames(submotifs))),
            peak_type=c("up", "down"),
            tf = c(cond1, cond1))
}
out <- do.call(rbind, out)
out$tf[out$tf=="EF1a-SPI1"] <- "SPI1"
out$tf[out$tf=="EF1a-KLF1"] <- "KLF1"
out <- out %>% as.data.frame %>% dplyr::mutate(pct_withmotif=npeak_contain_motif_of_interest/npeak)
write.table(out, paste0(outdir, "/atac_diffpeak_pctwithmotif_allcond.txt"), sep="\t", quote=F)

out <- read.table(paste0(outdir, "/atac_diffpeak_pctwithmotif_allcond.txt"), sep="\t")
out[out$peak_type=="down", "pct_withmotif"] <- -out[out$peak_type=="down", "pct_withmotif"]
ranked_tf <- cordf %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character
ranked_tf <- ranked_tf[ranked_tf != "tetCTCF"]
out$tf <- factor(out$tf, levels=ranked_tf)

# plot a stacked bar plot 
ggplot(out, aes(x=tf, y=pct_withmotif*100, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="% of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="% of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/atac_diffpeak_pctwithmotif_stacked_bar.pdf"), height=5, width=7)
```

### absolute # of peaks containing motifs
```{r}
# out["log10_npeak_contain_motif_of_interest"] <- log10(out["npeak_contain_motif_of_interest"])
out[out$peak_type=="down", "npeak_contain_motif_of_interest"] <- -out[out$peak_type=="down", "npeak_contain_motif_of_interest"]
# out[out$peak_type=="down", "log10_npeak_contain_motif_of_interest"] <- -out[out$peak_type=="down", "log10_npeak_contain_motif_of_interest"]

p1 <- ggplot(out, aes(x=tf, y=npeak_contain_motif_of_interest, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="# of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="# of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))
p1
ggsave(paste0(plotdir, "/atac_diffpeak_numwithmotif_stacked_bar.pdf"), height=5, width=7)
# # the log plot looks funky especially for small values, don't use
# p2 <- ggplot(out, aes(x=tf, y=log10_npeak_contain_motif_of_interest, fill=peak_type)) +
#   geom_bar(stat="identity") +
#   geom_hline(yintercept=0, color="black", linewidth=0.8) +
#   scale_fill_manual(values=c("#1465AC", "#B31B21")) +
#   labs(title="# of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
#        y="# of Differential Peaks", x="Transcription Factors") +
#   theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))


```


## Correlation at each dosage
```{r}
celltype <- "HEK293T"
cond2 <- "GFP_1"
log2fc.cutoff <- 0.58
padj.cutoff <- 0.05

atacdiff_dir <- "../../output/02-atac/04/deseq_data_Plate_TF_Dose/"
rnadiff_dir <- paste0("../../output/03-rna/01/tf_group/", celltype)

plotdir <- "../../plots/02-atac/05/tf_dose"
outdir <- "../../output/02-atac/05/tf_dose"

dir.create(plotdir, showWarnings = F, recursive = T)  
dir.create(outdir, showWarnings = F, recursive = T)

submeta <- meta[meta$CellType==celltype,]
cond1_ls <- unique(submeta$TF_Dose)
cond1_ls <- cond1_ls[cond1_ls != cond2]

cordf <- data.frame(row.names=cond1_ls, n_atac_diff_peaks=rep(0, length(cond1_ls)), n_rna_diff_genes=rep(0, length(cond1_ls)))

for (cond1 in cond1_ls){
  message(cond1)
  if (length(grep("EF1a|tet", cond1))==1){
    ataccond1 <- gsub("EF1a-", "", cond1) %>% gsub("tet-", "tet", .)
  } else{
    ataccond1 <- cond1
  }
  atacdiff <- fread(Sys.glob(paste0(atacdiff_dir, "/deseq_*", ataccond1, "_vs_*", cond2, "_filteredConsensus_full.tsv")))
  rnadiff <- readRDS(paste0(rnadiff_dir, "/deseq_dds_", gsub("-", ".", cond1), "_vs_", cond2, ".rds"))
  
  atacdiff_sig <- atacdiff %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  rnadiff_sig <- rnadiff %>% as.data.frame %>% dplyr::filter(padj<=padj.cutoff & abs(log2FoldChange)>=log2fc.cutoff)
  
  cordf[cond1, "n_atac_diff_peaks"] <- nrow(atacdiff_sig)
  cordf[cond1, "n_atac_diff_peaks_up"] <- nrow(atacdiff_sig %>% dplyr::filter(log2FoldChange>0))
  cordf[cond1, "n_atac_diff_peaks_down"] <- nrow(atacdiff_sig %>% dplyr::filter(log2FoldChange<0))
  cordf[cond1, "n_rna_diff_genes"] <- nrow(rnadiff_sig)
  cordf[cond1, "n_rna_diff_genes_up"] <- nrow(rnadiff_sig %>% dplyr::filter(log2FoldChange>0))
  cordf[cond1, "n_rna_diff_genes_down"] <- nrow(rnadiff_sig %>% dplyr::filter(log2FoldChange<0))
}


cordf$TF_Dose <- row.names(cordf)
cordf$TF <- str_split(cordf$TF_Dose, "_", simplify=T)[,1]
cordf[cordf$TF=="EF1a-KLF1", "TF"] <- "KLF1"
cordf[cordf$TF=="EF1a-SPI1", "TF"] <- "SPI1"
cordf[cordf$TF=="tet-CTCF", "TF"] <- "tetCTCF"
cordf$TF <- as.factor(cordf$TF)
BL_stallion <- colpal[levels(cordf$TF),]
cordf$Plasmid_Dose <- str_split(cordf$TF_Dose, "_", simplify=T)[,2] %>% as.character

write.table(cordf, paste0(outdir, "/atac_rna_ndiff_cor_tf_dose.txt"), sep="\t", quote=F)

p1 <- ggplot(cordf, aes(x=n_atac_diff_peaks, y=n_rna_diff_genes, col=TF, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) + ggrepel::geom_label_repel(show.legend = F)
p2 <- ggplot(cordf, aes(x=n_atac_diff_peaks, y=n_rna_diff_genes, col=Plasmid_Dose, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) 
p3 <- ggplot(cordf, aes(x=n_atac_diff_peaks, y=n_rna_diff_genes, col=TF, label=TF)) + geom_point(aes(shape=Plasmid_Dose)) + 
  theme_classic() + scale_color_manual(values=BL_stallion) 
p4 <- ggplot(cordf, aes(x=log10(n_atac_diff_peaks+1), y=log10(n_rna_diff_genes+1), col=TF, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) 
p5 <- ggplot(cordf[cordf$TF!="tetCTCF",], aes(x=log10(n_atac_diff_peaks+1), y=log10(n_rna_diff_genes+1), col=TF, label=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) 
p6 <- ggplot(cordf, aes(x=n_atac_diff_peaks, y=n_rna_diff_genes, col=TF)) + geom_point(size=0.3) + 
  theme_classic() + scale_color_manual(values=BL_stallion) + facet_wrap(~Plasmid_Dose) + theme(axis.text.x = element_text(angle = 90))
p7 <- ggplot(cordf, aes(x=n_atac_diff_peaks, y=n_rna_diff_genes, col=as.numeric(Plasmid_Dose))) + geom_point(size=0.3) + 
  theme_classic() + scale_color_viridis() + facet_wrap(~TF) + theme(axis.text.x = element_text(angle = 90))
p8 <- ggplot(cordf, aes(x=log10(n_atac_diff_peaks+1), y=log10(n_rna_diff_genes+1), col=as.numeric(Plasmid_Dose))) + geom_point(size=0.3) + 
  theme_classic() + scale_color_viridis() + facet_wrap(~TF) + theme(axis.text.x = element_text(angle = 90))


pdf(paste0(plotdir, "/atac_rna_ndiff_cor_tf_dose.pdf"), height=5, width=7)
p1
p2
p3
p4
p5
p6
p7
p8
dev.off()



```
```{r}
cordf <- read.table(paste0(outdir, "/atac_rna_ndiff_cor_tf_dose.txt"), sep="\t")
head(cordf)
```

### Plot a stacked bar plot to show TF dose diff peaks/genes
```{r}
df_long <- tidyr::gather(cordf, key="atac_diffpeak_type", value="count", n_atac_diff_peaks_up, n_atac_diff_peaks_down)
df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::filter(Plasmid_Dose=="1") %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
df_long$Plasmid_Dose <- factor(df_long$Plasmid_Dose, levels=c("1", "0.75", "0.5", "0.25", "0.05"))
df_long$log10count <- log10(df_long$count+1)

# sort TF Dose 
df_long$TF_Dose <- paste0(df_long$TF, "_", df_long$Plasmid_Dose)
TF_Dose_level <- lapply(levels(df_long$TF), function(tf){lapply(levels(df_long$Plasmid_Dose), function(n){paste0(tf, "_", n)}) %>% unlist}) %>% unlist
TF_Dose_level <- TF_Dose_level[TF_Dose_level %in% df_long$TF_Dose]
df_long$TF_Dose <- factor(df_long$TF_Dose, levels=TF_Dose_level)

# Adjust the 'down' values to be negative
df_long$count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"] <- -df_long$count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"]
df_long$log10count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"] <- -df_long$log10count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"]

# Plot using ggplot
ggplot(df_long, aes(x=TF_Dose, y=count, fill=atac_diffpeak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="ATAC Differential Peaks (Up and Down)",
       y="Number of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1, size=5))

ggsave(paste0(plotdir, "/atac_ndiff_cor_updown_stacked_bar.pdf"), height=5, width=10)

# log y 
ggplot(df_long, aes(x=TF_Dose, y=log10count, fill=atac_diffpeak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="ATAC Differential Peaks (Up and Down)",
       y="Log10(Number of Differential Peaks + 1)", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1, size=5))

ggsave(paste0(plotdir, "/atac_ndiff_cor_updown_stacked_bar_logged.pdf"), height=5, width=10)


# rna 
df_long <- tidyr::gather(cordf, key="rna_diffgene_type", value="count", n_rna_diff_genes_up, n_rna_diff_genes_down)
## sort by RNA diff genes at highest dose
# df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::filter(Plasmid_Dose=="1") %>% dplyr::arrange(desc(n_rna_diff_genes_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
# or sorted by ATAC diff peak num at highest dose
df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::filter(Plasmid_Dose=="1") %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
df_long$Plasmid_Dose <- factor(df_long$Plasmid_Dose, levels=c("1", "0.75", "0.5", "0.25", "0.05"))
df_long$log10count <- log10(df_long$count+1)

# sort TF Dose 
df_long$TF_Dose <- paste0(df_long$TF, "_", df_long$Plasmid_Dose)
TF_Dose_level <- lapply(levels(df_long$TF), function(tf){lapply(levels(df_long$Plasmid_Dose), function(n){paste0(tf, "_", n)}) %>% unlist}) %>% unlist
TF_Dose_level <- TF_Dose_level[TF_Dose_level %in% df_long$TF_Dose]
df_long$TF_Dose <- factor(df_long$TF_Dose, levels=TF_Dose_level)

# Adjust the 'down' values to be negative
df_long$count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"] <- -df_long$count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"]
df_long$log10count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"] <- -df_long$log10count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"]

# Plot using ggplot
ggplot(df_long, aes(x=TF_Dose, y=count, fill=rna_diffgene_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="RNA Differential Genes (Up and Down)",
       y="Number of Differential Genes", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1, size=5))

ggsave(paste0(plotdir, "/rna_ndiff_cor_updown_stacked_bar.pdf"), height=5, width=10)

# log y 
ggplot(df_long, aes(x=TF_Dose, y=log10count, fill=rna_diffgene_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="RNA Differential Genes (Up and Down)",
       y="Log10(Number of Differential Genes + 1)", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1, size=5))

ggsave(paste0(plotdir, "/rna_ndiff_cor_updown_stacked_bar_logged.pdf"), height=5, width=10)

```
### % of diff peaks up/down that contain motif of interest
```{r}

celltype <- "HEK293T"
cond2 <- "GFP_1"
log2fc.cutoff <- 0.58
padj.cutoff <- 0.05

atacdiff_dir <- "../../output/02-atac/04/deseq_data_Plate_TF_Dose/"

plotdir <- "../../plots/02-atac/05/tf_dose"
outdir <- "../../output/02-atac/05/tf_dose"

submeta <- meta[meta$CellType==celltype,]
cond1_ls <- unique(submeta$TF_Dose)
cond1_ls <- cond1_ls[cond1_ls != cond2]

cond1_ls <- cond1_ls[grep(paste0(names(motif_code_ls), collapse="|"), cond1_ls)]

out <- list()
for (cond1 in cond1_ls){
  message(cond1)
  if (length(grep("EF1a|tet", cond1))==1){
    ataccond1 <- gsub("EF1a-", "", cond1) %>% gsub("tet-", "tet", .)
  } else{
    ataccond1 <- cond1
  }
  submotifs <- motifs[,motif_code_ls[str_split(ataccond1, "_", simplify=T)[1]]]
  submotifs <- submotifs[motifMatches(submotifs) %>% as.list %>% unlist ==TRUE,]
  
  df <- fread(Sys.glob(paste0(atacdiff_dir, "/deseq_*", ataccond1, "_vs_*", cond2, "_filteredConsensus_full.tsv")))
  up <- df %>% dplyr::filter(padj<=padj.cutoff & log2FoldChange>=log2fc.cutoff)
  down <- df %>% dplyr::filter(padj<=padj.cutoff & log2FoldChange<=-log2fc.cutoff)
  out[[cond1]] <- data.frame(npeak=c(nrow(up), nrow(down)),
            npeak_contain_motif_of_interest=c(sum(up$name %in% rownames(submotifs)), sum(down$name %in% rownames(submotifs))),
            peak_type=c("up", "down"),
            tf_dose = c(cond1, cond1))
}

out <- do.call(rbind, out)
out <- out %>% as.data.frame %>% dplyr::mutate(pct_withmotif=npeak_contain_motif_of_interest/npeak)
write.table(out, paste0(outdir, "/atac_diffpeak_pctwithmotif_allcond.txt"), sep="\t", quote=F)

out[out$peak_type=="down", "pct_withmotif"] <- -out[out$peak_type=="down", "pct_withmotif"]

# sort TF Dose 
out$TF <- str_split(out$tf_dose, "_", simplify=T)[,1]
out[out$TF=="EF1a-KLF1", "TF"] <- "KLF1"
out[out$TF=="EF1a-SPI1", "TF"] <- "SPI1"
out[out$TF=="tet-CTCF", "TF"] <- "tetCTCF"

out$Plasmid_Dose <- factor(str_split(out$tf_dose, "_", simplify=T)[,2], levels=c("1", "0.75", "0.5", "0.25", "0.05"))
out$TF <- factor(out$TF, levels= out %>% dplyr::filter((Plasmid_Dose=="1") & (peak_type=="up")) %>% dplyr::arrange(desc(npeak)) %>% dplyr::select(TF) %>% unlist %>% as.character)
out$TF_Dose <- paste0(out$TF, "_", out$Plasmid_Dose)
TF_Dose_level <- lapply(levels(out$TF), function(tf){lapply(levels(out$Plasmid_Dose), function(n){paste0(tf, "_", n)}) %>% unlist}) %>% unlist
TF_Dose_level <- TF_Dose_level[TF_Dose_level %in% out$TF_Dose]
out$TF_Dose <- factor(out$TF_Dose, levels=TF_Dose_level)

# plot a stacked bar plot 
ggplot(out, aes(x=TF_Dose, y=pct_withmotif*100, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="% of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="% of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/atac_diffpeak_pctwithmotif_stacked_bar.pdf"), height=5, width=10)

```
### absolute # of peaks w/ motif
```{r}
out[out$peak_type=="down", "npeak_contain_motif_of_interest"] <- -out[out$peak_type=="down", "npeak_contain_motif_of_interest"]

p1 <- ggplot(out, aes(x=TF_Dose, y=npeak_contain_motif_of_interest, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="# of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="# of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))
p1
ggsave(paste0(plotdir, "/atac_diffpeak_numwithmotif_stacked_bar.pdf"), height=5, width=10)
```

## Highest dose only

```{r}
plotdir <- "../../plots/02-atac/05/tf_dose"
outdir <- "../../output/02-atac/05/tf_dose"

cordf <- read.table(paste0(outdir, "/atac_rna_ndiff_cor_tf_dose.txt"), sep="\t")
```

### stacked bar plot
```{r}
cordf <- cordf %>% dplyr::filter((Plasmid_Dose==1) & (TF!="tetCTCF"))

df_long <- tidyr::gather(cordf, key="atac_diffpeak_type", value="count", n_atac_diff_peaks_up, n_atac_diff_peaks_down)
df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
                                               
# Adjust the 'down' values to be negative
df_long$count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"] <- -df_long$count[df_long$atac_diffpeak_type == "n_atac_diff_peaks_down"]

# Plot using ggplot
ggplot(df_long, aes(x=TF, y=count, fill=atac_diffpeak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="ATAC Differential Peaks (Up and Down)",
       y="Number of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/atac_ndiff_cor_updown_stacked_bar_highestdose.pdf"), height=5, width=7)

# rna 
df_long <- tidyr::gather(cordf, key="rna_diffgene_type", value="count", n_rna_diff_genes_up, n_rna_diff_genes_down)
# df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::arrange(desc(n_rna_diff_genes_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
# or sorted by ATAC diff peak num at highest dose
df_long$TF <- factor(df_long$TF, levels= cordf %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character)
                                               
# Adjust the 'down' values to be negative
df_long$count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"] <- -df_long$count[df_long$rna_diffgene_type == "n_rna_diff_genes_down"]

# Plot using ggplot
ggplot(df_long, aes(x=TF, y=count, fill=rna_diffgene_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="RNA Differential Genes (Up and Down)",
       y="Number of Differential Genes", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/rna_ndiff_cor_updown_stacked_bar_highestdose.pdf"), height=5, width=7)
```

### % of diff peaks w/ motif of interest
```{r}
out <- read.table(paste0(outdir, "/atac_diffpeak_pctwithmotif_allcond.txt"), sep="\t")
out[c("tf", "dose")] <- str_split(out$tf_dose, "_", simplify=T)
out[out$tf=="EF1a-KLF1", "tf"] <- "KLF1"
out[out$tf=="EF1a-SPI1", "tf"] <- "SPI1"
out <- out %>% dplyr::filter(dose==1)

out[out$peak_type=="down", "pct_withmotif"] <- -out[out$peak_type=="down", "pct_withmotif"]
ranked_tf <- cordf %>% dplyr::arrange(desc(n_atac_diff_peaks_up)) %>% dplyr::select(TF) %>% unlist %>% as.character
ranked_tf <- ranked_tf[ranked_tf != "tetCTCF"]
out$tf <- factor(out$tf, levels=ranked_tf)

# plot a stacked bar plot 
ggplot(out, aes(x=tf, y=pct_withmotif*100, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="% of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="% of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))

ggsave(paste0(plotdir, "/atac_diffpeak_pctwithmotif_stacked_bar_highestdose.pdf"), height=5, width=7)
```

### absolute # of peaks w/ motif
```{r}
out[out$peak_type=="down", "npeak_contain_motif_of_interest"] <- -out[out$peak_type=="down", "npeak_contain_motif_of_interest"]

p1 <- ggplot(out, aes(x=tf, y=npeak_contain_motif_of_interest, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="# of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="# of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))
p1
ggsave(paste0(plotdir, "/atac_diffpeak_numwithmotif_stacked_bar_highestdose.pdf"), height=5, width=7)
```


also plot the absolute number of peaks containing motifs
```{r}
# out["log10_npeak_contain_motif_of_interest"] <- log10(out["npeak_contain_motif_of_interest"])
out[out$peak_type=="down", "npeak_contain_motif_of_interest"] <- -out[out$peak_type=="down", "npeak_contain_motif_of_interest"]
# out[out$peak_type=="down", "log10_npeak_contain_motif_of_interest"] <- -out[out$peak_type=="down", "log10_npeak_contain_motif_of_interest"]

p1 <- ggplot(out, aes(x=tf, y=npeak_contain_motif_of_interest, fill=peak_type)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept=0, color="black", linewidth=0.8) +
  scale_fill_manual(values=c("#1465AC", "#B31B21")) +
  labs(title="# of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
       y="# of Differential Peaks", x="Transcription Factors") +
  theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))
p1
ggsave(paste0(plotdir, "/atac_diffpeak_numwithmotif_stacked_bar.pdf"), height=5, width=7)
# # the log plot looks funky especially for small values, don't use
# p2 <- ggplot(out, aes(x=tf, y=log10_npeak_contain_motif_of_interest, fill=peak_type)) +
#   geom_bar(stat="identity") +
#   geom_hline(yintercept=0, color="black", linewidth=0.8) +
#   scale_fill_manual(values=c("#1465AC", "#B31B21")) +
#   labs(title="# of ATAC Differential Peaks Containing Motif of Interest (Up and Down)",
#        y="# of Differential Peaks", x="Transcription Factors") +
#   theme_classic() + theme(axis.text.x = element_text(angle=90, hjust=1))


```

